---
title: "Plotting the rise of Machine Learning health health related papers"
author: "Emmanuel Kagning Tsinda"
date: "October 13, 2018"
output: html_document
---

I would like to know how often biomedical and life science researchers are make use of machine learning algorithms. That's why I'm going to make a visual representation of the number of papers published in the field of medecine. Let's interrogate Pubmed, and look at the publication trend of machine learning scientific papers published in pubmed database. It seems that the 1st ever neural network was made in 1957 (https://www.forbes.com/sites/bernardmarr/2016/02/19/a-short-history-of-machine-learning-every-manager-should-read/#2b96042815e7) 



```{r}
library(rentrez)
library(ggplot2)
library(plotly)
search_year <- function(year, term){
  query <- paste(term, "AND (", year, "[PDAT])")
  entrez_search(db="pubmed", term=query, retmax=0)$count
}

year <- 1957:2014

```


Let's create a vector named papers
```{r}
papers <- sapply(year, search_year, term="machine learning", USE.NAMES=FALSE)

```


Let's build a dataframe containing the number of papers and year of publication 
```{r}
d = data.frame(year=year, number_of_papers=papers);d
```



Now, let's make an interactive plot displaying the number of papers and year of publication


```{r}

p=ggplot(data=d,aes(x=year,y=number_of_papers)) + geom_point(col="blue") + xlab("year") + ylab("number of published papers") + theme_bw() + theme(plot.title = element_text(hjust=0.5)) + 
  labs(title = "The rise of machine learning in biomedical and life sciences",
       caption = "based on data from NCBI/PUBMED", 
       subtitle= "based on data from NCBI/PUBMED",
       x = "year", y = "number of papers");ggplotly(p) 




```

Just like in other scientific disciplines, machine learning and AI is making progress. There are more and more health researchers who are trying to make use of the huge amount of healthcare data to predict or automate services. I can't wait to see the great medical discoveries !   

```{r}

plot_ly(x=~year, y=~number_of_papers, data=d, type="scatter", mode="marker", marker = list(color = "blue"))  %>%  
  layout(title = "The rise of machine learning in biomedical sciences and life sciences", 
    annotations = list(x = 1, y = -0.1, text = "based on data from NCBI/PUBMED", 
      showarrow = F, xref='paper', yref='paper',xanchor='right', yanchor='auto', xshift=0, yshift=0, font=list(size=10, color="blue"),xaxis = "x", yaxis = "y")
 )

```

There is a sharp increase of the number of publications from the year 2000. To better visualize that, lets build another plot

```{r}

plot_ly(x=~year, y=~number_of_papers, data=d[34:58, ], type="scatter", mode="marker", marker = list(color = "blue"), linetype = I("dash"))  %>%  
  layout(title = "The rise of machine learning in biomedical sciences and life sciences (1990-2017)", 
    annotations = list(x = 1, y = -0.1, text = "based on data from NCBI/PUBMED", 
      showarrow = F, xref='paper', yref='paper',xanchor='right', yanchor='auto', xshift=0, yshift=0, font=list(size=10, color="blue"),xaxis = "x", yaxis = "y")
 )


```

When I was trying to discuss this result, I realized Hashem et al, 2017 (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5760972/) found a quite similar trend, though he used a different R package (RISmed package) and plotted the proportion of papers per million. I think that the linear increase between 2005-2010 may be explained by the increased use of 3rd generation sequencing methods, which gave rise to big genomic data. Analysing such big data often requires dimension reduction and other machine learning methods.  Linear regression models were found to be the most dominant machine learning techniques in the life sciences over the past three decades 

Thank you for reading ! 
# Create a shareable link to the chart
# Set up API credentials: https://plot.ly/r/getting-started
chart_link = api_create(p, filename="text-mode")
chart_link
